Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-14 Hyperplane Enumeration/main.tex
More details at the extreme sizes, first pass of abstract

\begin{abstract}
TODO

@@

Here we introduce a method for finding all local optima of problems
with up to 80 bits. The cornerstone of this method is the identification
of hyperplanes that cannot contain any local optima, allowing large
portions of the search space to be skipped during enumeration.
The techniques used apply to the generalized problem class of Mk Landscapes,
which contains many real world and benchmark combinatorial problems.

@@

\subsection{Finding Local Optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
mean time to find all local optima all but 2 times. In both cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
As expected Gray-Box scales at almost exactly $2^N$. Using Hyperplane elimination without
reordering reduces this complexity, with reordering making further improvements. In the final
column we should the growth rate of the number of local optima. It is impossible for any method
which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
due to lower order terms dominating runtime for small problem sizes, as can be see in
Figure~\ref{fig-length-method}. For instance, fitting
the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
$m=0.358$ on Adjacent NKq.

With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
appears to be growing at the same rate as the total number of local optima. We suspect
some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
non-linear relationships per variable. Random NKq and MAX-kSAT both increase with problem
size, with Random NKq ranging between 5 and 6. With $N=15$, MAX-kSAT has on average nearly
12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
variable depend on over half the genome significantly limits the size of hyperplanes which
can be eliminated.

@@

\subsection{Finding $r$-Bit Local Optima}

>>>>

\begin{abstract}
A common tool for understanding combinatorial problem landscapes
is to examine the properties of its local optima. Unfortunately,
exhaustive methods for finding local optima are exceptionally time
consuming and are therefore limited to very small problem sizes.
We propose a method for exploiting generalized problem structure
to significantly speed up this process by skipping hyperplanes
which cannot contain local optima. In most of our experiments
this hyperplane elimination allows enumeration time to scale with
the total number of local optima instead of the total landscape size.
We prove optimality for linear and $k$-bounded problems, and provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
Furthermore, we are able to extend this method to finding only solutions
which cannot be improving by flipping $r$ or fewer bits. In some cases
increasing $r$ actually reduces total runtime by drastically reducing the
number of local optima. While previous methods were limited to
problems with at most 18 or 30 bits, hyperplane elimination can enumerate the same problems with
75 bits. When using higher $r$ values this bound can be pushed to 200 bit problems
on average enumerated in under 6 minutes.

@@

Here we introduce a method for finding all local optima for the same problems using
up to 75 bits. The cornerstone of this method is the identification
of hyperplanes that cannot contain any local optima, allowing large
portions of the search space to be skipped during enumeration.
The techniques used apply to the generalized problem class of Mk Landscapes,
which contains many real world and benchmark combinatorial problems.

@@

\subsection{Finding Local Optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
mean time to find all local optima all but 2 times. In both of those cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
As expected Gray-Box scales at almost exactly $2^N$. Using Hyperplane elimination without
reordering reduces this complexity, with reordering making further improvements. In the final
column we show the growth rate of the number of local optima. It is impossible for any method
which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
due to lower order terms dominating runtime for small problem sizes, as can be see in
Figure~\ref{fig-length-method}. For instance, fitting
the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
$m=0.358$ on Adjacent NKq. For Concatenated Traps this matches our theoretical predictions that
Hyper-Reorder scales at $2^k2^{N/k}$ as $\frac{1}{k}=0.2$.

This reduction in growth complexity translates into a significant increase in the size of problems
which can be enumerated. For instance, using the previous best enumeration methods~\cite{ochoa:2015:crossovernetworks}
the largest Adjacent NKq problem size which can be enumerated in 4 hours is 34 bits. Given the same time
Hyper-Reorder can enumerate problems with 75 bits. Using the regression model, we estimate solving
a 75 bit problem using Gray-Box would require 230 million years. Similarly, Hyper-Reorder's largest
Random NKq size of 69 bits would require Gray-Box 3 million years. These larger sizes make for
much better approximations of landscapes interesting to search methods.

@@

With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
appears to be growing at the same rate as the total number of local optima. We suspect
some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
non-linear relationships per variable. For small problem sizes
Random NKq and MAX-kSAT both increase with problem
size. Random NKq's average ranges between 5 and 6 on problems we tested.
With $N=15$, MAX-kSAT has on average nearly
12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
variable depend on over half the genome significantly limits the size of hyperplanes which
can be eliminated. While the number of non-linear relationships per variable
is asymptotically constant, this growth on small problems is likely affecting runtime.

\subsection{Finding $r$-Bit Local Optima}
A major advantage of Gray-Box enumeration is that it can find $r$-bit local optima
without exponentially increasing runtime complexity. Figure~\ref{fig-slope-method}
shows the estimated growth rate for each method as $r$ is increased. On Concatenated Traps,
Adjacent NKq, Ising Spin Glass, and to a lesser extent Random NKq, none of our enumeration
methods see an increase in growth complexity with increasing $r$. In all cases using Hyper-Reorder
has a lower growth rate than the other two methods using the same $r$. On every problem
except MAX-kSAT, Hyper-Reorder's slowest $r$ is faster than the fastest $r$
of any other configuration.

Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.

TODO Mention bigger problem sizes.
